Reanalysis Step 7: Missingness Pattern Analysis

### 2025-09-01 reanalysis-7-missingness_patterns.qmd
fix ai idiiocy

```{r variable-selection}
# Get variables used in outcome and propensity score models
# Based on the formulas in functions.R

# Core demographic variables
demographic_vars <- c("age", "age_cat", "gender_cat", "race_recode", "ethnicity_recode")

@@

# Combine all variables
all_vars <- c(demographic_vars, insurance_vars, diabetes_vars, 
              comorbidity_vars, vision_vars, exposure_var)

@@

if (length(missing_vars) > 0) {
  log_message("Variables not found in data: ", paste(missing_vars, collapse = ", "))
}

log_message("Analyzing ", length(available_vars), " variables")
```

@@

```{r descriptive-stats}
# Create Table 1 comparing excluded vs included
# Specify variable types
factor_vars <- c("age_cat", "gender_cat", "race_recode", "ethnicity_recode",
                 "insurance_cat", "insulin", "HTN", "dyslipidemia", "CVD",
                 "CKD_stage", "baseline_VA_cat", "dr_severity", "any_treatment",
                 "DME", "ever_lapse_binary")

@@

# Analyze lapsing rate
if ("ever_lapse_binary" %in% available_vars) {
  lapse_included <- mean(original_data$ever_lapse_binary[!original_data$excluded] == "Lapsed", na.rm = TRUE)
  lapse_excluded <- mean(original_data$ever_lapse_binary[original_data$excluded] == "Lapsed", na.rm = TRUE)
  lapse_p <- chisq.test(table(original_data$excluded, original_data$ever_lapse_binary))$p.value

@@

# Plot 2: Missingness by lapse status
if ("ever_lapse_binary" %in% available_vars) {
  miss_by_lapse <- original_data %>%
    group_by(ever_lapse_binary) %>%
    summarise(
      n_total = n(),
      n_excluded = sum(excluded),
      pct_excluded = 100 * n_excluded / n_total,
      .groups = 'drop'
    )

  plots[[2]] <- ggplot(miss_by_lapse, aes(x = ever_lapse_binary, y = pct_excluded)) +
    geom_col(fill = "darkgreen", alpha = 0.7) +
    geom_text(aes(label = sprintf("%.1f%%", pct_excluded)), 
              vjust = -0.5, size = 4) +
    labs(title = "Missingness by Lapse Status",
         x = "Lapse Status", y = "% Excluded") +
    theme_minimal()
}

# Plot 3: Joint distribution
if ("any_treatment" %in% available_vars && "ever_lapse_binary" %in% available_vars) {
  miss_joint <- original_data %>%
    group_by(any_treatment, ever_lapse_binary) %>%
    summarise(
      n_total = n(),
      n_excluded = sum(excluded),
      pct_excluded = 100 * n_excluded / n_total,
      .groups = 'drop'
    )

  plots[[3]] <- ggplot(miss_joint, aes(x = any_treatment, y = pct_excluded, 
                                       fill = ever_lapse_binary)) +
    geom_col(position = position_dodge(width = 0.8), alpha = 0.7) +
    labs(title = "Missingness by Treatment and Lapse Status",
         x = "Treatment Status", y = "% Excluded",
         fill = "Lapse Status") +
    theme_minimal() +
    scale_fill_manual(values = c("Not Lapsed" = "blue", "Lapsed" = "red"))
}

@@

# Calculate effect size measures
lapse_or_excluded_vs_included <- NA
if ("ever_lapse_binary" %in% available_vars) {
  # Calculate odds ratio of being excluded for lapsed vs not lapsed
  lapse_table <- table(original_data$ever_lapse_binary, original_data$excluded)
  if (ncol(lapse_table) == 2 && nrow(lapse_table) == 2) {
    lapse_or_excluded_vs_included <- (lapse_table[2,2] * lapse_table[1,1]) / 
                                     (lapse_table[2,1] * lapse_table[1,2])
  }
}

>>>>

```{r variable-selection}
# Get variables used in outcome and propensity score models
# First, check what columns exist in the original data
cat("Columns in original data:\n")
cat(paste(names(original_data)[1:30], collapse = "\n"))
cat("\n...\n")

# Based on the actual column names used in functions.R formulas
# Core demographic variables
demographic_vars <- c("age", "age_cat", "gender_cat", "race_ethnic_cat")

@@

# Outcome variables
outcome_vars <- c("outcome_va_logMAR", "outcome_va_vi_binary")

# Combine all variables
all_vars <- c(demographic_vars, insurance_vars, diabetes_vars, 
              comorbidity_vars, vision_vars, exposure_var, outcome_vars)

@@

if (length(missing_vars) > 0) {
  log_message(paste("Variables not found in data:", paste(missing_vars, collapse = ", ")))
}

log_message(paste("Analyzing", length(available_vars), "variables"))
```

@@

```{r descriptive-stats}
# Create Table 1 comparing excluded vs included
# Specify variable types (using actual column names from formulas)
factor_vars <- c("age_cat", "gender_cat", "race_ethnic_cat", "insurance_cat", 
                 "insulin", "HTN", "dyslipidemia", "CVD",
                 "CKD_stage", "baseline_VA_cat", "dr_severity", "any_treatment",
                 "DME", "ever_lapse_binary")

@@

# Analyze lapsing rate
if ("ever_lapse_binary" %in% available_vars) {
  lapse_included <- mean(original_data$ever_lapse_binary[!original_data$excluded] == 1, na.rm = TRUE)
  lapse_excluded <- mean(original_data$ever_lapse_binary[original_data$excluded] == 1, na.rm = TRUE)
  lapse_p <- chisq.test(table(original_data$excluded, original_data$ever_lapse_binary))$p.value

@@

# Plot 2: Missingness by lapse status
if ("ever_lapse_binary" %in% available_vars) {
  miss_by_lapse <- original_data %>%
    mutate(lapse_status = factor(ever_lapse_binary, 
                                 levels = c(0, 1),
                                 labels = c("Not Lapsed", "Lapsed"))) %>%
    group_by(lapse_status) %>%
    summarise(
      n_total = n(),
      n_excluded = sum(excluded),
      pct_excluded = 100 * n_excluded / n_total,
      .groups = 'drop'
    )

  plots[[2]] <- ggplot(miss_by_lapse, aes(x = lapse_status, y = pct_excluded)) +
    geom_col(fill = "darkgreen", alpha = 0.7) +
    geom_text(aes(label = sprintf("%.1f%%", pct_excluded)), 
              vjust = -0.5, size = 4) +
    labs(title = "Missingness by Lapse Status",
         x = "Lapse Status", y = "% Excluded") +
    theme_minimal()
}

# Plot 3: Joint distribution
if ("any_treatment" %in% available_vars && "ever_lapse_binary" %in% available_vars) {
  miss_joint <- original_data %>%
    mutate(lapse_status = factor(ever_lapse_binary, 
                                 levels = c(0, 1),
                                 labels = c("Not Lapsed", "Lapsed"))) %>%
    group_by(any_treatment, lapse_status) %>%
    summarise(
      n_total = n(),
      n_excluded = sum(excluded),
      pct_excluded = 100 * n_excluded / n_total,
      .groups = 'drop'
    )

  plots[[3]] <- ggplot(miss_joint, aes(x = any_treatment, y = pct_excluded, 
                                       fill = lapse_status)) +
    geom_col(position = position_dodge(width = 0.8), alpha = 0.7) +
    labs(title = "Missingness by Treatment and Lapse Status",
         x = "Treatment Status", y = "% Excluded",
         fill = "Lapse Status") +
    theme_minimal() +
    scale_fill_manual(values = c("Not Lapsed" = "blue", "Lapsed" = "red"))
}

@@

# Calculate effect size measures
lapse_or_excluded_vs_included <- NA
if ("ever_lapse_binary" %in% available_vars) {
  # Calculate odds ratio of being excluded for lapsed vs not lapsed
  lapse_table <- table(Lapse = original_data$ever_lapse_binary, 
                      Excluded = original_data$excluded)
  if (ncol(lapse_table) == 2 && nrow(lapse_table) == 2) {
    # OR = (lapsed & excluded * not lapsed & not excluded) / 
    #      (lapsed & not excluded * not lapsed & excluded)
    lapse_or_excluded_vs_included <- (lapse_table["1", "TRUE"] * lapse_table["0", "FALSE"]) / 
                                     (lapse_table["1", "FALSE"] * lapse_table["0", "TRUE"])
  }
}
